---
title: "ADS503 - Module 1 Lab"
author: "Satya Allumallu"
date: "1/20/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE)
```

### R Data Types

We will review vectors, matrices, data frames and list data types, these are the most commonly used data types in R. For more details on R data types see <https://www.statmethods.net/input/datatypes.html>. We will also see how to find more information about these R objects.


```{r datatypes}

xn <- c(1, 2, 3) # numeric vector
yn = c(1, 2, 3)

str(xn) # structure of the R object
class(xn) # class of the object

xc <- c('a', 'b', 'c') # character vector

str(xc)
class(xc)

xm <- matrix(1:20, nrow = 5, ncol = 4) # matrix
xm
?matrix

xm1 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) # matrix
xm1

str(xm1)
class(xm1)

xl <- list(a = 1:10, beta = exp(-3:3), logic = c(TRUE, FALSE, FALSE, TRUE)) # list
xl$a
xl[[2]]

str(xl)
class(xl)

xd <- data.frame(x = 1:10, y = letters[1:10], 
                 z = rep(c(TRUE, FALSE), 5)) # data frame
str(xd)
class(xd)
head(xd)

xd1 <- data.frame(x = 1:10, y = letters[1:10], 
                 z = rep(c(TRUE, FALSE), 5), stringsAsFactors = FALSE) # data frame
str(xd1)
head(xd1)
names(xd1) # names of columns in data frame
dim(xd1) # dimension of data frame
nrow(xd1) # number of rows of data frame
ncol(xd1) # number of columns of data frame

# Accessing data frames

xd1[1, ] # first row
xd1[, 1] # first column
xd1[2, 3] # second row and third column
xd1[, c("y", "z")] # data frame columns y and z
xd1[xd1$x > 5, ] # subset data frame
xds <- subset(xd1, x > 5) # subset data frame
subset(xd1, x > 5 & z == TRUE) # subset data frame
xds1 <- subset(xd1, x > 5, select = c(x, y)) # subset and select data frame
xds2 <- subset(xd1, x > 5, select = -z) # subset and select data frame

```


### Reading and writing files in R


```{r readwrite}

getwd()
setwd('/Users/sallumallu/Desktop/Online_Courses/USD_Course/MADS503/Module1')
read_df1 <- read.csv('/Users/sallumallu/Desktop/Online_Courses/USD_Course/MADS503/Module1/hcvdat0.csv')
str(read_df1)
read_df2 <- read.csv('/Users/sallumallu/Desktop/Online_Courses/USD_Course/MADS503/Module1/hcvdat0.csv',
                    stringsAsFactors = FALSE)
str(read_df2)

write.csv(xds2, file = 'xds2.csv')

```

### Apply functions in R.

We will review some of the apply functions in R. These are very useful and handy functions in base R that help you avoid using loops and write clean code. For more details on apply functions see <https://ademos.people.uic.edu/Chapter4.html>.

```{r applyfunctions}

data(cars)
head(cars)
dim(cars)

apply(cars, 1, mean) # row mean
apply(cars, 2, mean) # column mean

data(airquality)
head(airquality)
aq_class <- lapply(airquality, class) # get class of each column in the data frame
aq_class
aq_mean <- lapply(airquality, mean) # get mean of each column in the data frame
aq_mean
aq_mean_cleaned <- lapply(airquality, mean, na.rm = TRUE) # get mean of each column in the data frame
aq_mean_cleaned
airquality_s <- airquality[, lapply(airquality, class) == 'integer'] # subset
# data frame using lapply function

sapply(airquality, function(x) sum(is.na(x))) # using sapply to find NA's
aq_mean_cleaned_vector <- sapply(airquality, mean, na.rm = TRUE)
aq_mean_cleaned_vector
names(aq_mean_cleaned_vector)

```


### Some useful functions in R for module 1

pairs, cor, boxplot and skewness

```{r usefulfunctions}

data(iris)
str(iris)

pairs(iris) # pairwise scatter plots
#cor(iris) # correlation between predictors

cor(iris[,-5])

# some boxplots
boxplot(iris)

par(mfrow = c(1,2))
boxplot(iris$Sepal.Length ~ iris$Species)
boxplot(iris$Sepal.Width ~ iris$Species)

# some histograms
par(mfrow = c(1,3))
hist(iris$Sepal.Length)
hist(iris$Sepal.Width)
hist(iris$Petal.Length)

library(Hmisc)
hist.data.frame(iris)
hist.data.frame(airquality)

```

### Boxcox Transformations 

Boxcox transformations are used to transform a predictor variable with non-normal distribution into predictor with normal distribution.

```{r boxcox}

set.seed(20)
x <- rgamma(1000, 2)
hist(x)

library(e1071)
skewness(x)

y <- rnorm(1000)
hist(y)
skewness(y)

library(caret)
bct <- BoxCoxTrans(x)
x_trans <- predict(bct, x)
hist(x_trans)
skewness(x_trans)
         
z <- c(x, 0)
bct2 <- BoxCoxTrans(z)
z <- z + 10^-4
bct2 <- BoxCoxTrans(z)

```

### Principal Component Analysis

It is commonly used for dimensionality reduction by projecting each data point 
onto only the first few principal components to obtain lower-dimensional data 
while preserving as much of the data's variation as possible. 

Principal components are directions that have maximum variance when data is 
projected onto them and at the same time are uncorrelated with each other.

Principal components are linear combinations of original predictors.

```{r PCA}

# Example 1

set.seed(100)
x <- rnorm(1000)
y <- rnorm(1000)
z <- rnorm(1000)

m <- matrix(data = c(x, y, z), nrow = 1000, ncol = 3)

m_pca <- prcomp(m)
str(m_pca)

m_pca_var <- m_pca$sdev^2
m_pca_var * 100/sum(m_pca_var)
m_pca_var
m_pca$rotation
# m_pca$x # Use this to print all rows
m_pca$x[1:20,] # Use this to print first 20 rows

# Example 2

t <- (x + y + z)/3
m1 <- matrix(data = c(x, y, z, t), nrow = 1000, ncol = 4)

m1_pca <- prcomp(m1)
m1_pca_var <- m1_pca$sdev^2
m1_pca_var * 100/sum(m1_pca_var)

biplot(m1_pca, scale = 0)
biplot(m1_pca, choices = 2:3, scale = 0)


# Example 3

t <- exp(x)* y + z
m2 <- matrix(data = c(x, y, z, t), nrow = 1000, ncol = 4)

m2_pca <- prcomp(m2, scale. = TRUE)
m2_pca_var <- m2_pca$sdev^2
cumsum(m2_pca_var) * 100/sum(m2_pca_var) # top 3 components capture more 90% of variance

biplot(m2_pca, scale = 0)
biplot(m2_pca, choices = 2:3, scale = 0)

# Example 4

x <- rnorm(1000)
y <- rnorm(1000)
z <- rnorm(1000)
t <- rnorm(1000, mean = 0, sd = 4)

m3 <- matrix(data = c(x, y, z, t), nrow = 1000, ncol = 4)

m3_pca <- prcomp(m3)
m3_pca_var <- m3_pca$sdev^2
m3_pca_var * 100/sum(m3_pca_var)

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
